Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-08-30 analysis/paper/paper.Rmd
Updated PCA figures, raman section, and references

@@

@@

# Methods
All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that likely were made from the same material, however, for the smaller sites there still remains some uncertainties.  
Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020). PCA modelling was performed using the Prediktera Evince software (www.prediktera.com), which uses the NIPALS algorithm [@bertils19 pp.9-10].

@@

<br>

@@

<br>

```{r vis-nir-pca, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm part of the spectrum. Quartzite with a darker surface group to the right in the model, while white objects group in the bottom left quadrant."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/VNIR_PCA_p1_p2_center_20210818.jpg')
```

<br>

```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark and light groups in the 400 - 750 nm region"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/vis_nir_average_spectra_20210702.png')
```

<br>

@@

<br>

```{r nir-pca, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of light absorbance in 1 000 - 2 500 nm region"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/NIR_PCA_p1_p2_center_20210818.jpg')
```

<br>

## Raman
There are two distinctive features present in the Raman data. The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 300 cm-1 and 1 600 cm-1, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the D and G band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered carbon structures whilst the G band is caused by stretching in the carbon bonds, which is common to graphite-based materials [@vollebregt_etal12; @lee_etal21].  
The PCA model (fig.X) shows a clear trend along the first component (75.9 %) related to the above two features. The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. 
The 2nd component (7.82 %) mainly seems to differentiate the measurements that feature a lot of noise and objects that were difficult to sample. The same is true for the 3rd component, although it does not seem to be as susceptible to noise as the 2nd component. The 3rd component also separates out samples with irregular readings that could possibly be related to sampling issues on surfaces that were difficult to sample.

<br>

```{r raman-average, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* around 1 300 cm-1 (D band) and 1 600 cm-1 (G band)"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/raman_average_spectra_dark_20210823.png')
```

<br>

@@

The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 86.5%. The loadings of the three components can be seen in figure \@ref(fig:xrf-loadings). In the score plot of P1 and P2, two large clusters emerges (fig. \@ref(fig:xrf-score-plot-p1-p2)). The loadings for ^13^Al, ^18^Sr, ^19^K, ^20^Ca, ^40^Zr, ^81^Ti group together in the top right quadrant whilst ^12^Mg, ^14^Si, ^15^P, ^16^S, ^23^V, and ^56^Ba group in the bottom right. As seen in figure \@ref(fig:xrf-score-plot-p1-p2), ^26^Fe is the only element positioned on the left side of the model. Interestingly, the left cluster contains the samples with lower ^14^Si values and the right cluster the higher values. These clusters still remain if you were to exclude ^14^Si from the model, however, suggesting there are other elements of relevance to the clustering. The “low ^14^Si” cluster shows a trend along P2 of quartzite with a darker hue grouping towards the top. This may indicate a possible relation to the ^26^Fe content of the material, as a darker hue may be the result of inclusions of hematite, or possibly biotite. ^13^Al follows a similar trend with higher values recorded at the top of the two clusters. 
The *brecciated quartz* samples largely group together at the bottom of the “low ^14^Si” cluster, with a couple ending up towards the bottom of the “high ^14^Si” cluster. The white (N9) material is somewhat evenly distributed within the leftmost cluster. 
The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the top are all but one from the rock painting site of ‘Finnforsberget’, which is located next to the Skellefte river. The ‘Finnforsberget’ artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %. This artefact was sampled twice on each side, resulting in similar results both times.
One almost translucent sample also ends up among the darker material. Upon closer inspection this sample is extremely small, 3 by 1 mm, and also feature inclusions of dark particles. Both of these factors may have contributed to its placement in the model.

<br>

@@

<br>

@@

<br>

```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/XRF_PCA_p1_p2.png')
```

# Discussion
The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, none of them are capable of providing a detailed fingerprint on their own based solely on surface measurements. Taken together they do provide a more in-depth understanding of the characteristic and structure of the material. The most easily recognizable megascopic feature of the material is its colour, and the results of all three instrumentations demonstrate that there is some

>>>>

\linebreak

@@

\linebreak
 \linebreak

@@

\linebreak

# Methods

\linebreak
 \linebreak

All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that likely were made from the same material, however, for the smaller sites there still remains some uncertainties.  
Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020). PCA modelling was performed using the Prediktera Evince software (www.prediktera.com), which uses the NIPALS algorithm [@bertils19 pp.9-10].

\linebreak

@@

\linebreak
 \linebreak

@@

\linebreak

```{r vis-nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm region of the spectrum. Quartzite with a darker surface group to the right in the model, while white objects group in the bottom left quadrant."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_1_visnir_averaged_pca_p1_p2.jpg')
```

\linebreak

```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark and light groups in the 400 - 750 nm region. Dark material has a higher overall absorbance, with material closer to a white hue feature lower absorbance. The *translucent*/*opaque* group also feature higher absorbance in the ultraviolet region."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_2_average_visnir_spectra.jpg')
```

\linebreak

@@

\linebreak

```{r nir-spectra-comparison, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="Comparison of average spectra for dark, white, and translucent materials. Brecciated quartz is included as it is typically translucent with dark particle inclusion, as a comparison with the translucent group that mainly is made up of seemingly pure quarts."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/nir_spectra_comparison.jpg')
```

\linebreak

```{r nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant. A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
```

\linebreak

## Raman
There are two distinctive features present in the Raman data. The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the D and G band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered carbon structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being 1 330 - 1 350 cm-1. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca. 695 cm-1 and 795 cm-1. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on shale from XXXX. There is a fairly large band around 500 cm-1 which can be indicative of the presence of amorphous carbon (aC) [@hodkiewicz10], and it is likely the features observed above are another indication of carbon present within the samples. No other features were pr

The resulting PCA model (fig. \@ref(fig:nir-pca)) shows a clear trend along the first component (80.3 %) related to the above two features. The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. 
The 2nd component (6.82 %) mainly seems to differentiate spectra with a lot of noise, where the objects were difficult to sample. In contrast, along the 3rd component those samples that feature more pronounced peaks around 700 - 800 cm-1 groups up separately from the *quartz group*.

\linebreak

```{r raman-average, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_4_raman_spectra_dark.jpg')
```

\linebreak

```{r raman-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components of the raman analysis. The materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G band peaks around 1300 cm-1 and 1600 cm-1. Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak."}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
```

\linebreak

@@

The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 86.5%. The loadings of the three components can be seen in figure \@ref(fig:xrf-loadings). In the score plot of P1 and P2, two large clusters emerges (fig. \@ref(fig:xrf-score-plot-p1-p2)). The loadings for ^13^Al, ^18^Sr, ^19^K, ^20^Ca, ^40^Zr, ^81^Ti group together in the top right quadrant whilst ^12^Mg, ^14^Si, ^15^P, ^16^S, ^23^V, and ^56^Ba group in the bottom right. As seen in figure \@ref(fig:xrf-score-plot-p1-p2), ^26^Fe is the only element positioned on the left side of the model. Interestingly, the left cluster contains the samples with lower ^14^Si values and the right cluster the higher values. These clusters still remain if you were to exclude ^14^Si from the model, however, suggesting there are other elements of relevance to the clustering. The “low ^14^Si” cluster shows a trend along P2 of quartzite with a darker hue grouping towards the top. This may indicate a possible relation to the ^26^Fe content of the material, as a darker hue may be the result of inclusions of hematite, or possibly biotite. ^13^Al follows a similar trend with higher values recorded at the top of the two clusters. 
The *brecciated quartz* samples largely group together at the bottom of the “low ^14^Si” cluster, with a couple ending up towards the bottom of the “high ^14^Si” cluster. The white (N9) material is somewhat evenly distributed within the leftmost cluster. 
The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the top are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river. The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %. This artefact was sampled twice on each side, resulting in similar results both times.
One almost translucent sample also ends up among the darker material. Upon closer inspection this sample is extremely small, 3 by 1 mm, and also feature inclusions of dark particles. Both of these factors may have contributed to its placement in the model.

\linebreak

@@

\linebreak

@@

\linebreak

```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_7_xrf_pca_p1_p1.jpg')
```

\linebreak

# Discussion

\linebreak
 \linebreak

The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, none of them are capable of providing a detailed fingerprint on their own based solely on surface measurements. Taken together they do provide a more in-depth understanding of the characteristic and structure of the material. The most easily recognizable megascopic feature of the material is its colour, and the results of all three instrumentations demonstrate that there is some truth to this. However, a large group of the material, which features different hues of grey, does not conform as easily. In each of the PCA models there is a *greyish* middle that makes classification difficult.
